---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Panel data: One area without autocorrelation

## Aim

We are given a dataset containing daily counts of diseases from one geographical area. We want to identify:

- Does seasonality exist?
- If seasonality exists, when are the high/low seasons?
- Is there a general yearly trend (i.e. increasing or decreasing from year to year?)
- Is daily rainfall associated with the number of cases?

\newpage

## Creating the data

The data for this chapter is available at: http://rwhite.no/longitudinal_analysis/data/chapter_3.csv

```{r warning=FALSE}
# R CODE

dir.create("data")

library(data.table)
library(ggplot2)
set.seed(4)

AMPLITUDE <- 1.5
SEASONAL_HORIZONTAL_SHIFT <- 20

d <- data.table(date=seq.Date(
  from=as.Date("2000-01-01"),
  to=as.Date("2018-12-31"),
  by=1))
d[,year:=as.numeric(format.Date(date,"%G"))]
d[,week:=as.numeric(format.Date(date,"%V"))]
d[,month:=as.numeric(format.Date(date,"%m"))]
d[,yearMinus2000:=year-2000]
d[,dailyrainfall:=runif(.N, min=0, max=10)]

d[,dayOfYear:=as.numeric(format.Date(date,"%j"))]
d[,seasonalEffect:=sin(2*pi*(dayOfYear-SEASONAL_HORIZONTAL_SHIFT)/365)]
d[,mu := exp(0.1 + yearMinus2000*0.1 + seasonalEffect*AMPLITUDE)]
d[,y:=rpois(.N,mu)]

fwrite(d,"data/chapter_3.csv")
```

\newpage

## True data

Here we show the true data, and note that there is an increasing annual trend (the data gets higher as time goes on) and there is a seasonal pattern (one peak/trough per year)
```{r}
q <- ggplot(d,aes(x=date))
q <- q + geom_point(mapping=aes(y=y))
q <- q + geom_line(mapping=aes(y=mu),colour="red")
q
```

\newpage

## Investigation 

Pretending we have no prior knowledge of our dataset, we display the data for few years and see a clear seasonal trend
```{r}
q <- ggplot(d[year %in% c(2005:2010)],aes(x=dayOfYear,y=y))
q <- q + facet_wrap(~year)
q <- q + geom_point()
q <- q + stat_smooth(colour="red")
q
```

\newpage

## Seasonality 

If we want to investigate the seasonality of our data, and identify when are the peaks and troughs, we have a few ways to approach this.

Non-parametric approaches are flexible and easy to implement, but they can lack power and be hard to interpret:

- Create a categorical variable for the seasons (e.g. `spring`, `summer`, `autumn`, `winter`) and include this in the regression model
- Create a categorical variable for the months (e.g. `Jan`, `Feb`, ..., `Dec`) and include this in the regression model

Parametric approaches are more powerful but require more effort:

- Identify the periodicity of the seasonality (how many days between peaks?)
- Using trigonometry, transform `day of year` into variables that appropriately model the observed periodicity
- Obtain coefficient estimates
- Back-transform these estimates into human-understandable values (day of peak, day of trough)

The non-parametric approaches are simple and we will therefore not cover them in this course. We will briefly examine the parametric approach.

*NOTE:* You don't always have to investigate seasonality! It depends entirely on what the purpose of your analysis is!

\newpage

The Lomb-Scargle Periodogram shows a clear seasonality with a period of 365 days.

```{}
// STATA CODE STARTS
insheet using "chapter_3.csv", clear

sort date
gen time=_n
tsset time, daily

wntestb y

cumsp y, gen(cumulative_spec_dist)
gen period=_N/_n

browse cumulative_spec_dist period
// STATA CODE ENDS
```

```{r}
# R CODE 
lomb::lsp(d$y,from=100,to=500,ofac=1,type="period")
```

\newpage

We then generate two new variables `cos365` and `sin365` and perform a likelihood ratio test to see if they are significant or not. This is done with two simple poisson regressions.

When we do not have autocorrelation, we can use the `glm` function in R and in STATA. Note that it is very important to specify the `family` (as this is how we differentiate between linear/logistic/poisson regressions).
```{}
// STATA CODE STARTS
gen cos365=cos(dayofyear*2*_pi/365)
gen sin365=sin(dayofyear*2*_pi/365)

glm y yearminus2000 dailyrainfall, family(poisson)
estimates store m1
glm y yearminus2000 dailyrainfall cos365 sin365, family(poisson)
estimates store m2

predict resid, anscombe

lrtest m1 m2
// STATA CODE ENDS
```

```{r}
# R CODE
d[,cos365:=cos(dayOfYear*2*pi/365)]
d[,sin365:=sin(dayOfYear*2*pi/365)]

fit0 <- glm(y~yearMinus2000 + dailyrainfall, data=d, family=poisson())
fit1 <- glm(y~yearMinus2000 + dailyrainfall + sin365 + cos365, data=d, family=poisson())

print(lmtest::lrtest(fit0, fit1))
```

We see that the likelihood ratio test for `sin365` and `cos365` was significant, meaning that there is significant seasonality with a 365 day periodicity in our data (which we already strongly suspected due to the periodogram).

\newpage

We can now run/look at the results of our main regression.

```{r}
print(summary(fit1))
```
We also see that the (significant!) coefficient for `year` is `0.1` which means that for each additional year, the outcome increases by `exp(0.1)=1.11`. We also see that the coefficient for `dailyrainfall` was not significant, which means that we did not find a significant association between the outcome and `dailyrainfall`.

*NOTE:* See that this is basically the same as a normal regression.

\newpage

Through the likelihood ratio test we saw a clear significant seasonal effect. We can now use trigonometry to back-calculate the amplitude and location of peak/troughs from the `cos365` and `sin365` estimates:
```{r}
b1 <- 1.428417 # sin coefficient
b2 <- -0.512912 # cos coefficient
amplitude <- sqrt(b1^2 + b2^2)
p <- atan(b1/b2) * 365/2/pi
if (p > 0) {
    peak <- p
    trough <- p + 365/2
} else {
    peak <- p + 365/2
    trough <- p + 365
}
if (b1 < 0) {
    g <- peak
    peak <- trough
    trough <- g
}
print(sprintf("amplitude is estimated as %s, peak is estimated as %s, trough is estimated as %s",round(amplitude,2),round(peak),round(trough)))

print(sprintf("true values are: amplitude: %s, peak: %s, trough: %s",round(AMPLITUDE,2),round(365/4+SEASONAL_HORIZONTAL_SHIFT),round(3*365/4+SEASONAL_HORIZONTAL_SHIFT)))

```

*NOTE:* An amplitude of 1.5 means that when comparing the average time of year to the peak, the peak is expected to be `exp(1.5)=4.5` times higher than average. We take the exponential because we have run a poisson regression (so think incident rate ratio).

\newpage

We now investigate our residuals to determine if we have a good fit:

```{r}
d[,residuals:=residuals(fit1, type = "response")]
d[,predicted:=predict(fit1, type = "response")]
q <- ggplot(d,aes(x=predicted,y=residuals))
q <- q + geom_point()
q <- q + stat_smooth(colour="red")
q
```

\newpage

We check the `pacf` of the residuals to ensure that it is not `AR`. If we observe `AR` in our residuals, then this model was not appropriate and we need to use a different model.

```{}
// STATA CODE STARTS
pac resid
// STATA CODE ENDS
```

```{r}
# R CODE
# this is for AR
pacf(d$residuals)
```

\newpage

We check the `acf` of the residuals to ensure that it is not `MA`. If we observe `MA` in our residuals, then this model was not appropriate and we need to use a different model.

```{}
// STATA CODE STARTS
ac resid
// STATA CODE ENDS
```

```{r}
# R CODE
# this is for MA
acf(d$residuals)
```
