---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Panel data: One area with autocorrelation

## Aim

We are given a dataset containing daily counts of diseases from one geographical area. We want to identify:

- Does seasonality exist?
- If seasonality exists, when are the high/low seasons?
- Is there a general yearly trend (i.e. increasing or decreasing from year to year?)

(We remove the question about rainfall in order to simplify and streamline the exercise)

\newpage

## Creating the data

The data for this chapter is available at: http://rwhite.no/longitudinal_analysis/data/chapter_4.csv

```{r}
library(data.table)
library(ggplot2)
set.seed(4)

AMPLITUDE <- 1.5
SEASONAL_HORIZONTAL_SHIFT <- 20

d <- data.table(date=seq.Date(
  from=as.Date("2000-01-01"),
  to=as.Date("2018-12-31"),
  by=1))
d[,year:=as.numeric(format.Date(date,"%G"))]
d[,week:=as.numeric(format.Date(date,"%V"))]
d[,month:=as.numeric(format.Date(date,"%m"))]
d[,yearMinus2000:=year-2000]
d[,dayOfSeries:=1:.N]

d[,dayOfYear:=as.numeric(format.Date(date,"%j"))]
d[,seasonalEffect:=sin(2*pi*(dayOfYear-SEASONAL_HORIZONTAL_SHIFT)/365)]
d[,mu := exp(0.1 + yearMinus2000*0.1 + seasonalEffect*AMPLITUDE)]
d[,y:=rpois(.N,mu)]
d[,y:=round(as.numeric(arima.sim(model=list("ar"=c(0.5)), rand.gen = rpois, n=nrow(d), lambda=mu)))]

fwrite(d,"data/chapter_4.csv")
```

\newpage

## Investigation 

We display the data for few years and see a clear seasonal trend
```{r}
q <- ggplot(d[year %in% c(2005:2010)],aes(x=dayOfYear,y=y))
q <- q + facet_wrap(~year)
q <- q + geom_point()
q <- q + stat_smooth(colour="red")
q
```

\newpage

The Lomb-Scargle Periodogram shows a clear seasonality with a period of 365 days

```{}
// STATA CODE STARTS
insheet using "chapter_4.csv", clear

sort date
gen time=_n
tsset time, daily

wntestb y

cumsp y, gen(cumulative_spec_dist)
gen period=_N/_n

browse cumulative_spec_dist period
// STATA CODE ENDS
```

```{r}
# R CODE
lomb::lsp(d$y,from=50,to=500,ofac=1,type="period")
```

\newpage

## Regressions

We then generate two new variables `cos365` and `sin365` and perform a likelihood ratio test to see if they are significant or not. This is done with two simple poisson regressions.
```{}
// STATA CODE STARTS
gen cos365=cos(dayofyear*2*_pi/365)
gen sin365=sin(dayofyear*2*_pi/365)

glm y yearminus2000, family(poisson)
estimates store m1
glm y yearminus2000 cos365 sin365, family(poisson)
estimates store m2

predict resid, anscombe

lrtest m1 m2
// STATA CODE ENDS
```

```{r}
# R CODE
d[,cos365:=cos(dayOfYear*2*pi/365)]
d[,sin365:=sin(dayOfYear*2*pi/365)]

fit0 <- glm(y~yearMinus2000, data=d, family=poisson())
fit1 <- glm(y~yearMinus2000+sin365 + cos365, data=d, family=poisson())

print(lmtest::lrtest(fit0, fit1))
```

We see that the likelihood ratio test for `sin365` and `cos365` was significant, meaning that there is significant seasonality with a 365 day periodicity in our data (which we already strongly suspected due to the periodogram).

\newpage

We can now run/look at the results of our main regression.

```{r}
print(summary(fit1))
```
We also see that the coefficient for year is `0.1` which means that for each additional year, the outcome increases by `exp(0.1)=1.11`. 

\newpage

## Residual analysis

```{r}
d[,residuals:=residuals(fit1, type = "response")]
d[,predicted:=predict(fit1, type = "response")]
```

We can see a clear `AR(1)` pattern in our residuals.
```{}
// STATA CODE STARTS
pac resid
// STATA CODE ENDS
```

```{r}
# R CODE
# this is for AR
pacf(d$residuals)
```

\newpage

And again we see some sort of `AR` pattern in our residuals.
```{}
// STATA CODE STARTS
ac resid
// STATA CODE ENDS
```

```{r}
# R CODE
# this is for MA
acf(d$residuals)
```

This means our model is bad, we have autocorrelation. We now need to change our model to account for this `AR(1)` autocorrelation!

\newpage

## (R ONLY) Regression with AR(1) correlation in residuals

First we create an `id` variable. This generally corresponds to geographical locations, or people. In this case, we only have one geographical location, so our `id` for all observations is `1`. This lets the computer know that all data belongs to the same group.

When we have autocorrelation in the residuals, we can use the `MASS::glmPQL` function in R.
```{r}
d[,ID:=1]
# this is for MA
fit <- MASS::glmmPQL(y~yearMinus2000+sin365 + cos365, random = ~ 1 | ID,
                family = poisson, data = d,
                correlation=nlme::corAR1(form=~dayOfSeries|ID))
summary(fit)
```

\newpage 

We can see that the residuals no longer display any signs of autocorrelation.
```{r}
pacf(residuals(fit, type = "normalized")) # this is for AR
```

\newpage

We can see that the residuals no longer display any signs of autocorrelation.
```{r}
acf(residuals(fit, type = "normalized")) # this is for MA
```

\newpage

We also obtain the same estimates that we did in the last chapter.

```{r}
b1 <- 1.3936185 # sin coefficient
b2 <- -0.5233866 # cos coefficient
amplitude <- sqrt(b1^2 + b2^2)
p <- atan(b1/b2) * 365/2/pi
if (p > 0) {
    peak <- p
    trough <- p + 365/2
} else {
    peak <- p + 365/2
    trough <- p + 365
}
if (b1 < 0) {
    g <- peak
    peak <- trough
    trough <- g
}
print(sprintf("amplitude is estimated as %s, peak is estimated as %s, trough is estimated as %s",round(amplitude,2),round(peak),round(trough)))

print(sprintf("true values are: amplitude: %s, peak: %s, trough: %s",round(AMPLITUDE,2),round(365/4+SEASONAL_HORIZONTAL_SHIFT),round(3*365/4+SEASONAL_HORIZONTAL_SHIFT)))

```

\newpage

## (STATA ONLY) Regression with robust standard errors

In STATA it is not possible to explicitly model autocorrelation in the residuals (with the exception of linear regression). Since most of our work deals with logistic and poisson regressions, we will be focusing on modelling strategies that work with all kinds of regressions.

The STATA approach to autocorrelation is to estimate more `robust` standard errors. That is, STATA makes the standard errors larger to account for the model mispecification. This is done through the `vce(robust)` option.

```{}
// STATA CODE STARTS
glm y yearminus2000 cos365 sin365, family(poisson) vce(robust)
// STATA CODE ENDS
```

