---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Not panel data: Multiple areas

## Aim

We are given a dataset containing counts of diseases from multiple geographical areas. We want to identify:

- Is there a general yearly trend (i.e. increasing or decreasing from year to year?)
- Is variable `x` associated with the outcome?

\newpage

## Creating the data

The data for this chapter is available at: http://rwhite.no/longitudinal_analysis/data/chapter_5.csv

```{r warning=FALSE}
library(data.table)
library(lme4)
set.seed(4)

fylkeIntercepts <- data.table(fylke=1:20,fylkeIntercepts=rnorm(20))

d <- data.table(fylke=rep(1:20,each=100))
d <- merge(d,fylkeIntercepts,by="fylke")
d[,mainIntercept:=3]
d[,x:=runif(.N)]
d[,year:=sample(c(1950:2018),.N,replace=T)]
d[,mu := exp(mainIntercept + fylkeIntercepts + 3*x)]
d[,y:=rpois(.N,mu)]

fwrite(d,"data/chapter_5.csv")
```

\newpage

## Investigating the data

We can see from the data that we have 20 geographical areas (`fylke`) with 100 observations for each fylke, but the sampling did not happen consistently (some years have multiple measurements, other years have no measurements).

This means we have:

- multiple geographical areas
- multiple observations in each geographical area
- not panel data

```{r}
print(d)
```

\newpage

## Regression

For this scenario, we use the `lme4::glmer` function in R. We need to introduce a `(1|fylke)` term to identify the geographical areas (i.e. clusters). In STATA we use the `meglm` function and introduce a `|| fylke:` term to identify the geographical areas (i.e. clusters).

```{}
// STATA CODE STARTS
insheet using "chapter_5.csv", clear

gen yearMinus2000 = year-2000
meglm y x yearMinus2000 || fylke:, family(poisson)
// STATA CODE ENDS
```

```{r warning=FALSE}
# R CODE
d[,yearMinus2000:=year-2000]
summary(fit <- lme4::glmer(y~x + yearMinus2000 + (1|fylke),data=d,family=poisson()))
```

You can see that the format of the results is the same as an ordinary regression.


