---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Panel data: multiple areas without autocorrelation

## Aim

We are given a dataset containing daily counts of diseases from multiple geographical areas. We want to identify:

- Does seasonality exist?
- If seasonality exists, when are the high/low seasons?
- Is there a general yearly trend (i.e. increasing or decreasing from year to year?)

\newpage

## Creating the data

The data for this chapter is available at: http://rwhite.no/longitudinal_analysis/data/chapter_6.csv

```{r}
library(data.table)
library(ggplot2)
set.seed(4)

AMPLITUDE <- 1.5
SEASONAL_HORIZONTAL_SHIFT <- 20

fylkeIntercepts <- data.table(fylke=1:20,fylkeIntercepts=rnorm(20))

d <- data.table(date=seq.Date(
  from=as.Date("2010-01-01"),
  to=as.Date("2015-12-31"),
  by=1))
d[,year:=as.numeric(format.Date(date,"%G"))]
d[,week:=as.numeric(format.Date(date,"%V"))]
d[,month:=as.numeric(format.Date(date,"%m"))]

temp <- vector("list",length=20)
for(i in 1:20){
  temp[[i]] <- copy(d)
  temp[[i]][,fylke:=i]
}
d <- rbindlist(temp)

d[,yearMinus2000:=year-2000]
d[,dayOfSeries:=1:.N]

d[,dayOfYear:=as.numeric(format.Date(date,"%j"))]
d[,seasonalEffect:=sin(2*pi*(dayOfYear-SEASONAL_HORIZONTAL_SHIFT)/365)]
d[,mu := exp(0.1 + yearMinus2000*0.1 + seasonalEffect*AMPLITUDE)]
d[,y:=rpois(.N,mu)]

fwrite(d,"data/chapter_6.csv")
```

\newpage

## Investigation
We then drill down into a few years for fylke 1, and see a clear seasonal trend
```{r warning=FALSE}
q <- ggplot(d[fylke==1],aes(x=dayOfYear,y=y))
q <- q + facet_wrap(~year)
q <- q + geom_point()
q <- q + stat_smooth(colour="red")
q
```

\newpage

The Lomb-Scargle Periodogram shows a clear seasonality with a period of 365 days

```{}
// STATA CODE STARTS
insheet using "chapter_6.csv", clear

sort fylke date
by fylke: gen time=_n
tsset fylke time, daily

wntestb y if fylke==1

cumsp y if fylke==1, gen(cumulative_spec_dist)
by fylke: gen period=_N/_n

browse cumulative_spec_dist period
// STATA CODE ENDS
```

```{r}
# RCODE
lomb::lsp(d$y,from=100,to=500,ofac=1,type="period")
```

\newpage

## Regression

First we create an `id` variable. This generally corresponds to geographical locations, or people. In this case, we only have one geographical location, so our `id` for all observations is `1`. This lets the computer know that all data belongs to the same group.

When we have panel data with multiple areas, we use the `MASS::glmPQL` function in R and the `meglm` function in STATA. In R we identify the geographical areas with `random = ~ 1 | fylke` and in STATA with `|| fylke:`.
```{}
// STATA CODE STARTS
gen cos365=cos(dayofyear*2*_pi/365)
gen sin365=sin(dayofyear*2*_pi/365)

meglm y yearminus2000 || fylke:, family(poisson) iter(10)
estimates store m1
meglm y yearminus2000 cos365 sin365 || fylke:, family(poisson) iter(10)
estimates store m2

predict resid, anscombe

lrtest m1 m2
// STATA CODE ENDS
```

```{r}
# R CODE
d[,cos365:=cos(dayOfYear*2*pi/365)]
d[,sin365:=sin(dayOfYear*2*pi/365)]
fit0 <- MASS::glmmPQL(y~yearMinus2000, random = ~ 1 | fylke,
                family = poisson, data = d)
fit1 <- MASS::glmmPQL(y~yearMinus2000 + sin365 + cos365, random = ~ 1 | fylke,
                family = poisson, data = d)

print(lmtest::lrtest(fit0, fit1))
```

We see that the likelihood ratio test for `sin365` and `cos365` was significant, meaning that there is significant seasonality with a 365 day periodicity in our data (which we already strongly suspected due to the periodogram).

\newpage

We can now run/look at the results of our main regression.

```{r}
print(summary(fit1))
```

\newpage

## Residual analysis

We see that there is no evidence of autoregression in the residuals
```{}
// STATA CODE STARTS
pac resid if fylke==1
// STATA CODE ENDS
```

```{r}
# R CODE
pacf(residuals(fit1, type = "normalized")) # this is for AR
```

\newpage

We see that there is no evidence of autoregression in the residuals
```{}
// STATA CODE STARTS
ac resid if fylke==1
// STATA CODE ENDS
```

```{r}
# R CODE
acf(residuals(fit1, type = "normalized")) # this is for MA
```

\newpage

We also obtain the same estimates that we did in the last chapter.

```{r}
b1 <- 1.4007640 # sin coefficient
b2 <- -0.5234863 # cos coefficient
amplitude <- sqrt(b1^2 + b2^2)
p <- atan(b1/b2) * 365/2/pi
if (p > 0) {
    peak <- p
    trough <- p + 365/2
} else {
    peak <- p + 365/2
    trough <- p + 365
}
if (b1 < 0) {
    g <- peak
    peak <- trough
    trough <- g
}
print(sprintf("amplitude is estimated as %s, peak is estimated as %s, trough is estimated as %s",round(amplitude,2),round(peak),round(trough)))

print(sprintf("true values are: amplitude: %s, peak: %s, trough: %s",round(AMPLITUDE,2),round(365/4+SEASONAL_HORIZONTAL_SHIFT),round(3*365/4+SEASONAL_HORIZONTAL_SHIFT)))

```

