---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Panel data: multiple areas with autocorrelation

```{r}
library(data.table)
library(ggplot2)
set.seed(4)

AMPLITUDE <- 1.5
SEASONAL_HORIZONTAL_SHIFT <- 20

fylkeIntercepts <- data.table(fylke=1:20,fylkeIntercepts=rnorm(20))

d <- data.table(date=seq.Date(
  from=as.Date("2010-01-01"),
  to=as.Date("2015-12-31"),
  by=1))
d[,year:=as.numeric(format.Date(date,"%G"))]
d[,week:=as.numeric(format.Date(date,"%V"))]
d[,month:=as.numeric(format.Date(date,"%m"))]

temp <- vector("list",length=20)
for(i in 1:20){
  temp[[i]] <- copy(d)
  temp[[i]][,fylke:=i]
}
d <- rbindlist(temp)

d[,yearMinus2000:=year-2000]
d[,dayOfSeries:=1:.N]

d[,dayOfYear:=as.numeric(format.Date(date,"%j"))]
d[,seasonalEffect:=sin(2*pi*(dayOfYear-SEASONAL_HORIZONTAL_SHIFT)/365)]
d[,mu := round(exp(0.1 + yearMinus2000*0.1 + seasonalEffect*AMPLITUDE))]
d[,y:=rpois(.N,mu)]
d[,y:=mu+round(as.numeric(arima.sim(model=list("ar"=c(0.5)), rand.gen = rpois, n=nrow(d), lambda=mu)))]

```

\newpage

## Investigation

We drill down into a few years in fylke 1, and see a clear seasonal trend
```{r warning=FALSE}
q <- ggplot(d[fylke==1],aes(x=dayOfYear,y=y))
q <- q + facet_wrap(~year)
q <- q + geom_point()
q <- q + stat_smooth(colour="red")
q
```

\newpage

The Lomb-Scargle Periodogram shows a clear seasonality with a period of 365 days
```{r}
lomb::lsp(d$y,from=100,to=500,ofac=1,type="period")
```

\newpage

## Regressions
```{r}
d[,cos365:=cos(dayOfYear*2*pi/365)]
d[,sin365:=sin(dayOfYear*2*pi/365)]

fit <- MASS::glmmPQL(y~yearMinus2000+sin365 + cos365, random = ~ 1 | fylke,
                family = poisson, data = d)
summary(fit)
```

\newpage

## Residual analysis

We see that there is an `AR(1)` autocorrelation in the residuals, meaning that our model is not appropriate.
```{r}
pacf(residuals(fit, type = "normalized")) # this is for AR
```

\newpage

We see that there is some sort of `AR` autocorrelation in the residuals, meaning that our model is not appropriate.
```{r}
acf(residuals(fit, type = "normalized")) # this is for MA
```

\newpage

## Regression with AR(1) correlation in residuals

We include `correlation=nlme::corAR1(form=~dayOfSeries|fylke)` or in other words `correlation=nlme::corAR1(form=~time|group)` to let the computer know what is the time variable and what is the group variable.

```{r}
fit <- MASS::glmmPQL(y~yearMinus2000+sin365 + cos365, random = ~ 1 | fylke,
                family = poisson, data = d,
                correlation=nlme::corAR1(form=~dayOfSeries|fylke))
summary(fit)
```

\newpage

## Residual analysis

We see that the vast majority of the autoregression in the residuals has been removed.
```{r}
pacf(residuals(fit, type = "normalized")) # this is for AR
```

\newpage

We see that the vast majority of the autoregression in the residuals has been removed.
```{r}
acf(residuals(fit, type = "normalized")) # this is for MA
```

\newpage

We obtain the same estimates that we did in the last chapter.

```{r}
b1 <- 1.4007640 # sin coefficient
b2 <- -0.5234863 # cos coefficient
amplitude <- sqrt(b1^2 + b2^2)
p <- atan(b1/b2) * 365/2/pi
if (p > 0) {
    peak <- p
    trough <- p + 365/2
} else {
    peak <- p + 365/2
    trough <- p + 365
}
if (b1 < 0) {
    g <- peak
    peak <- trough
    trough <- g
}
print(sprintf("amplitude is estimated as %s, peak is estimated as %s, trough is estimated as %s",round(amplitude,2),round(peak),round(trough)))

print(sprintf("true values are: amplitude: %s, peak: %s, trough: %s",round(AMPLITUDE,2),round(365/4+SEASONAL_HORIZONTAL_SHIFT),round(3*365/4+SEASONAL_HORIZONTAL_SHIFT)))

```

